---
title: "Compare_PAVE"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
vignette: "%\\VignetteIndexEntry{TRACE_Demo} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Data set

Example LC-MS data from PAVE: [Google Drive download](https://drive.google.com/open?id=1C7gwniCSwdk2Povc0KpA_BBLbuCkzMRo)

```{r}

raw.data.dir <- "C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/raw_neg/"
dir(raw.data.dir)
```

### 1.1 Load data to MSdev object

```{r}
library(TRACE)
library(MSdev)
#object <- MSdev(rawDataDir = raw.data.dir)
#object <- MSdev_msConvert(object)

object <- MSdev_load("C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/MSdev_2026_06_03.Rdata")

object@sampleInfo
```

### 1.2 Set sample information and parameters

```{r}
#object <- MSdev_checkSampleInfo(object)
object <- MSdev_set_param(
    object,
    findChromPeaks =
      xcms::CentWaveParam(
        ppm = 5,
        prefilter = c(3,1000),
        peakwidth = c(10,30),
        snthresh = 10,
        fitgauss = T),
    groupChromPeaks =
      xcms::PeakDensityParam(
        sampleGroups = "A",
        minFraction = 0.6,
        binSize = 0.002,
        bw = 5,
        ppm = 5))
```

### 1.3 Extract peaks and group peaks using xcms

```{r}
#object <- MSdev_xcmsProcessing(object)
#object@xcmsData

MSdev_save(object)

```

### 1.4 TRACE analysis

re-network assignment
``` {r}
object <- TRACE_network_assignment(object, i.pol = 0)
object <- TRACE_annotate(object, i.pol = 0)
```

```{r}
#object <- TRACE_workflow(object)
TRACE_export(object, file  = "c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/TRACE_export.xlsx")
MSdev_save(object)
```

## 2. Result compare

### 2.1 Merge data

```{r}
{
  library(dplyr)

  pave.file <- "c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/pks_features.xlsx"
  trace.file <- "c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/TRACE_export.xlsx"


  pave.df.raw <- openxlsx::read.xlsx(pave.file)
  trace.df.raw <-  openxlsx::read.xlsx(  trace.file,sheet = 1 )
  pave.df <- pave.df.raw %>%
    dplyr::mutate(
      feature_id = as.integer(gsub("[^0-9]+", "", feature_id)),
      formula_pave = paste0("C", C_num, "N", N_num),
      type_pave = dplyr::case_when(
        is.na(feature) | feature == "Background" ~ "blank/noise",
        TRUE ~ as.character(feature)
      ),
      type_pave = case_when(
        type_pave %in% c("Metabolite","Isotope","Fragment","Adduct")~type_pave,
        T~"blank/noise"
      ),
      score_pave = score,
      pave_cn = formula_pave != "C0N0"
    ) %>%
    dplyr::select(
      feature_id,
      mz_pave = mz,
      rt_pave = rt,
      formula_pave,
      type_pave,
      score_pave,
      pave_cn,
      pave_seed,
      isotope
    )

  trace.df <- trace.df.raw %>%
    dplyr::mutate(
      feature_id = as.integer(feature_id),
      formula_trace = as.character(TRACE_formula),
      type_trace = dplyr::case_when(
        is.na(type) | type == "" ~ "blank/noise",
        TRUE ~ as.character(type)
      ),
      trace_cn =  !is.na(TRACE_formula),
      seed_trace = seed
    ) %>%
    dplyr::select(
      feature_id,
      mz_trace = mz,
      rt_trace = rt,
      formula_trace,
      type_trace,
      seed_trace,
      trace_cn,
      TRACE_cor,
      TRACE_net_score,
      TRACE_anno_score,
      adduct = compound.adduct,
      TRACE_fragment,
      TRACE_isotope
    )

 

  library(stringr)
  pave.trace.merged <- dplyr::full_join(pave.df, trace.df, by = "feature_id") %>%
    dplyr::mutate(
      type_pave = if_else(is.na(type_pave),"blank/noise",type_pave),
      type_pave = str_to_sentence(type_pave),
      type_trace = str_to_sentence(type_trace),
      cn_match = trace_cn &
        pave_cn &
        !is.na(formula_trace) &
        formula_trace != "" &
        formula_pave == formula_trace,
      cn_assignment = dplyr::case_when(
        trace_cn & pave_cn ~ "both",
        trace_cn & !pave_cn ~ "TRACE_only",
        !trace_cn & pave_cn ~ "PAVE_only",
        TRUE ~ "neither"
      )
    )




  pave.trace.merged <- pave.trace.merged %>%
    dplyr::select(
      feature_id,
      formula_pave,
      formula_trace,
      type_pave,
      type_trace,
      everything()
    ) %>%
    dplyr::arrange(
      type_pave,
      type_trace)




}
```

### 2.2 Plot annotation change

#### 2.2.1 Annotation change summary

```{r}
count.df <- pave.trace.merged %>%
  dplyr::count(type_trace, type_pave, name = "n")

p <- ggplot2::ggplot(pave.trace.merged, ggplot2::aes(x = type_trace, y = type_pave)) +
  ggplot2::geom_count(show.legend = FALSE) +
  ggplot2::scale_size(range = c(2, 13),
             transform = "sqrt") +
  ggplot2::geom_text(
    data = count.df,
    ggplot2::aes(x = type_trace, y = type_pave, label = n),
    color = "white",
    size = 2,
    show.legend = FALSE
  ) +
  ggplot2::labs(x = "TRACE annotation", y = "PAVE annotation") +
  ggplot2::theme_bw(base_size = 6)

open_plot_win(p, 2.2, 2)
```

#### 2.2.2 CN formula matched ratio

compare the CN formula matched ratio of TRACE and PAVE, the ratio is the number of features annotated as CN formula in TRACE divided by the number of features annotated as CN formula in PAVE. only metabolite and adduct features are considered. use a stack bar plot like : \|PAVE\|OVERLAP\|TRACE\|

```{r}
res <- .compare_pave_cn_match_ratio_plot(
  pave.trace.merged,
  types = c("Metabolite", "Adduct")#,"Fragment","Isotope")
)
open_plot_win(res$plot, 1.5, 2)
```

#### 2.2.3 False positive

show the proportion of false positive of CN formula, the false positive is the features annotated as CN formula in PAVE, but filtered out in TRACE by dynamic filter.

```{r}
res.cn.fp <- .compare_pave_cn_false_positive_plot(pave.trace.merged)
open_plot_win(res.cn.fp$plot, 1.5, 2)
```

show the proportion of false positive of adduct/fragment/isotope, the false positive is the features annotated as adduct/fragment/isotope in PAVE, but filtered out or re-assigned in TRACE by network assignment.

```{r}
res.afis.fp <- .compare_pave_afis_false_positive_plot(pave.trace.merged)
open_plot_win(res.afis.fp$plot, 1.5, 2)
```

### 2.3 Annotation change composition

#### 2.3.1 CN dynamic filter

pick all features annoated as non-blank in PAVE , but annotated as blank in TRACE, markerd as unmatched_blank. pick all features annotated as non-blank in both TRACE and PAVE, markerd as matched_cn. extract all unmatched_blank and matched_cn's CN bundles from MSdev object: 1. the bundle with seed as unmatched_blank/matched_cn should be extracted 2. the mz error in each bundle should be extracted, and the max mz error should be calculated 3. add arg 'mz_error', could be 'max' or 'mean', to control the mz error to be used and then plot: 1. boxplot of the CN bundles' cor of unmatched_blank and matched_cn 2. boxplot of the CN bundles' max mz error of unmatched_blank and matched_cn

```{r}
p <- plot_compare_pave_cn_bundle(pave.trace.merged, object, i.pol = 0,mz_error = "max")
open_plot_win(p, 5, 2.5)
```

#### 2.3.2 Adduct selection

PAVE select M+/-H as metabolites and others as adduct, but TRACE pick the max_intensity feature in a CN group as the seed and annotate it with the formula, and then annotate other features in the same CN group as adducts. So the adduct selection is different between these two methods. We can compare the adduct selection by comparing the adduct annotation of features in the same CN group.

extract all these features' information:

1.  features groupped as M/M, M/A, A/M, A/A in PAVE and TRACE
2.  intensity of each features

plot:

1.  barplot to show adduct type distribution of TRACE and PAVE, all of PAVE as M-H, and TRACE use the trace annotation. plot in one figure, two columns, one for TRACE, one for PAVE, the type of adduct should stocked in the bar.
2.  box plot to show the four group's intensity distribution in TRACE and PAVE, x is the group, y is the intensity, color is the group, the x axis title should be 'PAVE/TRACE'

```{r}
p <- plot_compare_pave_adduct(pave.trace.merged, object, i.pol = 0)
open_plot_win(p, 5, 2.5)
```

#### 2.3.3 Fragment compare

show the features annotated as fragment in PAVE, but not in TRACE,

1.  merge `pave_fragment >> trace_adduct` and `pave_fragment >> trace_metabolite` into one data.frame; one circular barplot (`geom_bar()` + `coord_polar()`) to show the adduct type in TRACE

2.  for pave_fragment \>\> trace_metabolite, circular barplot (`geom_bar()` + `coord_polar()`) to show whether the feature was a candidate fragment in network assignment (check if `feature_id` is in `fg.net`). the legend title should be 'Fragment candidate', fill should be 'Re-assigned candidate' and 'Non candidate'.

```{r}
p <- plot_compare_pave_fragment(pave.trace.merged, object, i.pol = 0)
open_plot_win(p, 5, 2.5)
```

#### 2.3.4 Graph compare

Plot the same cached network-split demo twice: TRACE colors nodes by
`assign.seed` (network-assignment group); PAVE colors by metabolite seed m/z.
Each `pave_seed` is matched to the closest m/z in `pave.trace.merged` to
obtain a seed `feature_id`; nodes without `pave_seed` are labeled `noise`
(grey). Layout `(x, y)` is computed once with `compute_TRACE_network_layout()`
and reused so the two panels are directly comparable. Edges and node borders are drawn in
grey with partial transparency so formula / type / m/z labels stay readable.
Edge relationship labels (adduct / fragment / isotope) are shown on the TRACE
panel.

```{r}
demo <- TRACE_load_network_split_demo(
  demo_id = 13,
  cache_dir = "C:/Users/91879/OneDrive/Code/R/Package/TRACE/cache"
)
layout_xy <- compute_TRACE_network_layout(demo$ig)

vertex_data <- demo$vertex_data %>%
  dplyr::mutate(
    feature_id = as.character(name),
    trace_seed = as.character(assign.seed)
  )

lab.df <- vertex_data %>%
  dplyr::left_join(
    pave.trace.merged %>%
      dplyr::mutate(feature_id = as.character(feature_id)) %>%
      dplyr::select(
        feature_id,
        formula_pave, type_pave, mz_pave,
        formula_trace, type_trace, mz_trace,
        pave_seed
      ),
    by = "feature_id"
  )

lab.df$pave_group <- map_pave_seed_vertex_group(
  feature_id = lab.df$feature_id,
  pave_seed = lab.df$pave_seed,
  pave.trace.merged = pave.trace.merged,
  type_pave = lab.df$type_pave
)

p1 <- plot_TRACE_network_split_demo(
  demo,
  plot_title = "TRACE network assignment",
  plot_subtitle = "",
  layout_coords = layout_xy,
  muted_style = TRUE,
  show_legend = TRUE,
  legend_title = "TRACE seed",
  show_edge_label = TRUE,
  vertex_group = stats::setNames(lab.df$trace_seed, lab.df$feature_id),
  vertex_label_data = data.frame(
    feature_id = lab.df$feature_id,
    formula = lab.df$formula_trace,
    type = lab.df$type_trace,
    mz = lab.df$mz_trace,
    stringsAsFactors = FALSE
  )
)

p2 <- plot_TRACE_network_split_demo(
  demo,
  plot_title = "PAVE metabolite grouping",
  plot_subtitle = "",
  layout_coords = layout_xy,
  muted_style = TRUE,
  show_legend = TRUE,
  legend_title = "PAVE seed ",
  show_edge_label = FALSE,
  vertex_group = lab.df$pave_group,
  vertex_label_data = data.frame(
    feature_id = lab.df$feature_id,
    formula = lab.df$formula_pave,
    type = lab.df$type_pave,
    mz = lab.df$mz_pave,
    stringsAsFactors = FALSE
  )
)

open_plot_win(
  p1 / p2 + patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "right"),
  6,
  10
)
```

```{r}
library(tidyr)
library(ggplot2)

edge.df <- as.data.frame(demo$edge_data)
ppm.by.vertex <- dplyr::bind_rows(
  edge.df %>% dplyr::transmute(feature_id = as.character(from), mz_ppm = abs(as.numeric(mz.ppm))),
  edge.df %>% dplyr::transmute(feature_id = as.character(to),   mz_ppm = abs(as.numeric(mz.ppm)))
) %>%
  dplyr::group_by(feature_id) %>%
  dplyr::summarise(ppm = stats::median(mz_ppm, na.rm = TRUE), .groups = "drop")

bar.df <- vertex_data %>%
  dplyr::transmute(
    feature_id = as.character(feature_id),
    trace_seed = as.character(trace_seed),
    rt = as.numeric(rt)
  ) %>%
  dplyr::left_join(ppm.by.vertex, by = "feature_id") %>%
  dplyr::arrange(rt) %>%
  dplyr::mutate(feature_id = factor(feature_id, levels = feature_id))

bar.long <- bar.df %>%
  tidyr::pivot_longer(
    cols = c(rt, ppm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    metric = factor(
      metric,
      levels = c("rt", "ppm"),
      labels = c("RT (s)", "m/z error (ppm)")
    )
  )

p <- ggplot2::ggplot(bar.long, ggplot2::aes(x = feature_id, y = value, col = trace_seed)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_grid(metric ~ ., scales = "free_y") +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    fill = "TRACE seed"
  ) +
  ggplot2::theme_bw(base_size = 6) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

open_plot_win(p, 2, 1.5)


```
