---
title: "Compare_PAVE"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
vignette: "%\\VignetteIndexEntry{TRACE_Demo} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Data set

Example LC-MS data from PAVE: [Google Drive download](https://drive.google.com/open?id=1C7gwniCSwdk2Povc0KpA_BBLbuCkzMRo)

```{r}

raw.data.dir <- "C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/raw_neg/"
dir(raw.data.dir)
```

### Load data to MSdev object

```{r}
library(TRACE)
library(MSdev)
#object <- MSdev(rawDataDir = raw.data.dir)
#object <- MSdev_msConvert(object)

object <- MSdev_load("C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/MSdev_2026_06_03.Rdata")

object@sampleInfo
```

### Set sample information and parameters

```{r}
#object <- MSdev_checkSampleInfo(object)
object <- MSdev_set_param(
    object,
    findChromPeaks =
      xcms::CentWaveParam(
        ppm = 5,
        prefilter = c(3,1000),
        peakwidth = c(10,30),
        snthresh = 10,
        fitgauss = T),
    groupChromPeaks =
      xcms::PeakDensityParam(
        sampleGroups = "A",
        minFraction = 0.6,
        binSize = 0.002,
        bw = 5,
        ppm = 5))
```

### Extract peaks and group peaks using xcms

```{r}
#object <- MSdev_xcmsProcessing(object)
#object@xcmsData

MSdev_save(object)
```

### TRACE analysis

```{r}
object <- TRACE_workflow(object)
MSdev_save(object)
```

### Compare PAVE and TRACE result
1. the PAVE-matlab result was stored in the `PAVE_matlab` folder, the TRACE result was stored in the `"c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/pks_features.xlsx"`
2. the TRACE result was stored in the `"d:/temp/TRACE.xlsx"`
3. read in both of these files and write code to compare the result, code append to "test/test.R" and placed in {}
4. the code should compare and plot:
  - a dot plot, x with annotation type of TRACE, y is annotation type of PAVE-matlab, the color is the correlation score of the TRACE result, the size is the number of features
  - a radar plot, the ratio of the CN formula matched
5. do not use mz/rt match, the feature_id has been matched in the TRACE_workflow, just use the feature_id to compare the result
6. the p1 should contain blank/noise
7. P2 use the pie plot, just show ratio of CN mathced/unmatched count of TRACE and PAVE-matlab
8. align the two result with feature_id: trace$feature_id  and pave$feature_id (its numeric part), do not use .match_mz_rt