---
title: "Compare_PAVE"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
vignette: "%\\VignetteIndexEntry{TRACE_Demo} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Data set

Example LC-MS data from PAVE: [Google Drive download](https://drive.google.com/open?id=1C7gwniCSwdk2Povc0KpA_BBLbuCkzMRo)

```{r}

raw.data.dir <- "C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/raw_neg/"
dir(raw.data.dir)
```

### Load data to MSdev object

```{r}
library(TRACE)
library(MSdev)
#object <- MSdev(rawDataDir = raw.data.dir)
#object <- MSdev_msConvert(object)

object <- MSdev_load("C:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/raw_neg-20251010T060553Z-1-001/MSdev_2026_06_03.Rdata")

object@sampleInfo
```

### Set sample information and parameters

```{r}
#object <- MSdev_checkSampleInfo(object)
object <- MSdev_set_param(
    object,
    findChromPeaks =
      xcms::CentWaveParam(
        ppm = 5,
        prefilter = c(3,1000),
        peakwidth = c(10,30),
        snthresh = 10,
        fitgauss = T),
    groupChromPeaks =
      xcms::PeakDensityParam(
        sampleGroups = "A",
        minFraction = 0.6,
        binSize = 0.002,
        bw = 5,
        ppm = 5))
```

### Extract peaks and group peaks using xcms

```{r}
#object <- MSdev_xcmsProcessing(object)
#object@xcmsData

MSdev_save(object)
TRACE_export(object, file  = "c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/TRACE_export.xlsx")
```

### TRACE analysis

```{r}
#object <- TRACE_workflow(object)
#MSdev_save(object)
```

### result 
```{r}
{
  library(dplyr)

  pave.file <- "c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/pks_features.xlsx"
  trace.file <- "d:/temp/TRACE.xlsx"
  trace.sheet <- "AllFeatures_Negative"

  pave.df.raw <- openxlsx::read.xlsx(pave.file)
  trace.df.raw <-  openxlsx::read.xlsx(  trace.file,sheet = trace.sheet )
  pave.df <- pave.df.raw %>%
    dplyr::mutate(
      feature_id = as.integer(gsub("[^0-9]+", "", feature_id)),
      formula_pave = paste0("C", C_num, "N", N_num),
      type_pave = dplyr::case_when(
        is.na(feature) | feature == "Background" ~ "blank/noise",
        TRUE ~ as.character(feature)
      ),
      type_pave = case_when(
        type_pave %in% c("Metabolite","Isotope","Fragment","Adduct")~type_pave,
        T~"blank/noise"
      ),
      score_pave = score,
      pave_cn = formula_pave != "C0N0"
    ) %>%
    dplyr::select(
      feature_id,
      mz_pave = mz,
      rt_pave = rt,
      formula_pave,
      type_pave,
      score_pave,
      pave_cn
    )

  trace.df <- openxlsx::read.xlsx(
    trace.file,
    sheet = trace.sheet
  ) %>%
    dplyr::mutate(
      feature_id = as.integer(feature_id),
      formula_trace = as.character(TRACE_formula),
      type_trace = dplyr::case_when(
        is.na(type) | type == "" ~ "blank/noise",
        TRUE ~ as.character(type)
      ),
      trace_cn =  !is.na(TRACE_formula),
      seed_trace = seed
    ) %>%
    dplyr::select(
      feature_id,
      mz_trace = mz,
      rt_trace = rt,
      formula_trace,
      type_trace,
      seed_trace,
      trace_cn
    )

  ratio.df <- tryCatch(
    openxlsx::read.xlsx(
      trace.file
    ),
    error = function(e) NULL
  )

  pave.trace.merged <- dplyr::full_join(pave.df, trace.df, by = "feature_id") %>%
    dplyr::mutate(
      type_pave = if_else(is.na(type_pave),"blank/noise",type_pave),
      cn_match = trace_cn &
        pave_cn &
        !is.na(formula_trace) &
        formula_trace != "" &
        formula_pave == formula_trace,
      cn_assignment = dplyr::case_when(
        trace_cn & pave_cn ~ "both",
        trace_cn & !pave_cn ~ "TRACE_only",
        !trace_cn & pave_cn ~ "PAVE_only",
        TRUE ~ "neither"
      )
    )

  if (!is.null(ratio.df) && all(c("TRACE_seed", "TRACE_cor") %in% names(ratio.df))) {
    trace.cor.map <- stats::setNames(ratio.df$TRACE_cor, as.character(ratio.df$TRACE_seed))
    pave.trace.merged$trace_cor <- as.numeric(
      trace.cor.map[as.character(pave.trace.merged$seed_trace)]
    )
  } else {
    pave.trace.merged$trace_cor <- NA_real_
  }
  pave.trace.merged$trace_cor <- ifelse(
    is.finite(pave.trace.merged$trace_cor),
    pave.trace.merged$trace_cor,
    pave.trace.merged$score_pave
  )

  pave.trace.merged <- pave.trace.merged %>%
    dplyr::select(
      feature_id,
      formula_pave,
      formula_trace,
      type_pave,
      type_trace,
      score_pave,
      trace_cor,
      pave_cn,
      trace_cn,
      cn_match,
      cn_assignment,
      mz_pave,
      rt_pave,
      mz_trace,
      rt_trace,
      seed_trace
    ) %>%
    dplyr::arrange(feature_id)


  ggplot(pave.trace.merged,aes(x = type_trace,y = type_pave))+
    geom_count()+
    scale_size(range = c(1,20),
               transform = "sqrt")+
    stat_sum(
      aes(label = after_stat(n)),
      geom = "text",
      color = "white",
      size = 3.5,
      show.legend = FALSE
    )


}
```




### Compare PAVE and TRACE result

1.  the PAVE-matlab result was stored in the `PAVE_matlab` folder, the TRACE result was stored in the `"c:/Users/91879/OneDrive/Documents/YLF_Lab/Project/2025.10.10.PAVE/PAVE-Matlab/example/pks_features.xlsx"`
2.  the TRACE result was stored in the `"d:/temp/TRACE.xlsx"`
3.  read in both of these files and write code to compare the result, code append to "test/test.R" and placed in {}
4.  the code should compare and plot:

- a dot plot, x with annotation type of TRACE, y is annotation type of PAVE-matlab, the color is the correlation score of the TRACE result, the size is the number of features
- a radar plot, the ratio of the CN formula matched

5.  do not use mz/rt match, the feature_id has been matched in the TRACE_workflow, just use the feature_id to compare the result
6.  the p1 should contain blank/noise
7.  P2 use the pie plot, just show ratio of CN mathced/unmatched count of TRACE and PAVE-matlab
8.  align the two result with feature_id: trace$feature_id  and pave$feature_id (its numeric part), do not use .match_mz_rt

#### CN net compare

1.  distinct the features assigned to CN from TRACE and PAVE-matlab, plot the ratio of features assigned to CN in TRACE but not in PAVE-matlab, and the ratio of features assigned to CN in PAVE-matlab but not in TRACE.
2.  merge the two result to one data.frame, with columns:

- feature_id, formula_pave, formula_trace, ......
