---
title: "cor_sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cor_sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TRACE)
library(MSdev)
library(dplyr)
library(ggplot2)
```


### 1. Scitific problem

 The authors should provide a sensitivity analysis across multiple ρ thresholds (TRACE_cor) and report effects on bundle numbers.





### 2. data set

Example LC-MS demo data (same as `TRACE_Demo`):
```{r}
trace.demo <- MSdev_load("d:/data/PAVE_demo/MSdev_2026_06_01.Rdata")
trace.demo@sampleInfo
```

`trace.demo` is used read-only below. Do not call `MSdev_save(trace.demo)` so the on-disk demo file is not overwritten.

### 3. evaluate

We perform a sensitivity analysis across multiple ρ thresholds (`TRACE_cor`), `seq(0.5, 0.95, 0.05)`, using the CN-finder table already stored in `trace.demo@advancedAna$TRACE_temp` (read-only). For each cutoff we count CN bundle edges (`cn.net.hit` rows) and CN seeds (unique bundle `from` IDs), then quantify seed-set overlap across thresholds.

#### 3.1 Bundle numbers across TRACE_cor cutoffs

```{r}
{
  cor.list <- seq(0.5, 0.95, 0.05)
  cor.labels <- paste0("rho=", cor.list)

  bundle.list <- lapply(c("Negative", "Positive"), function(pol) {
    cn.finder <- trace.demo@advancedAna$TRACE_temp[[pol]]$CNfinder
    if (is.null(cn.finder) || !nrow(cn.finder)) {
      return(NULL)
    }

    cn.finder <- data.table::as.data.table(cn.finder)

    lapply(seq_along(cor.list), function(i) {
      cn.hit <- cn.finder[TRACE_cor >= cor.list[i] & is.finite(TRACE_cor)]
      data.frame(
        cor.cut = cor.list[i],
        param.label = cor.labels[i],
        polarity = pol,
        n_edge = nrow(cn.hit),
        n_seed = length(unique(cn.hit$from)),
        mean_bundle_size = if (nrow(cn.hit)) {
          mean(table(cn.hit$from))
        } else {
          NA_real_
        },
        stringsAsFactors = FALSE
      )
    }) %>% dplyr::bind_rows()
  })

  bundle.summary <- dplyr::bind_rows(bundle.list) %>%
    dplyr::mutate(
      cor.cut = factor(cor.cut, levels = cor.list),
      param.label = factor(param.label, levels = cor.labels)
    )

  p.bundle <- ggplot2::ggplot(
    bundle.summary,
    ggplot2::aes(x = cor.cut, y = n_seed, color = polarity, group = polarity)
  ) +
    ggplot2::geom_line(linewidth = 0.6) +
    ggplot2::geom_point(size = 1.2) +
    ggplot2::scale_color_brewer(palette = "Set1", name = "Polarity") +
    ggplot2::labs(
      x = expression(rho ~ " "),
      y = "CN seed count",
      #title = "CN seeds retained across TRACE_cor cutoffs"
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

  p.edge <- ggplot2::ggplot(
    bundle.summary,
    ggplot2::aes(x = cor.cut, y = n_edge, color = polarity, group = polarity)
  ) +
    ggplot2::geom_line(linewidth = 0.6) +
    ggplot2::geom_point(size = 1.2) +
    ggplot2::scale_color_brewer(palette = "Set1", name = "Polarity") +
    ggplot2::labs(
      x = expression(rho ~ " "),
      y = "CN bundle edge count",
     # title = "CN bundle edges retained across TRACE_cor cutoffs"
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

  p.bundle.all <- p.bundle + p.edge +
    patchwork::plot_annotation(tag_levels = "A")+
    patchwork::plot_layout(guides = "collect")

  print(bundle.summary)
}

if (interactive()) {
  open_plot_win(p.bundle.all, 5, 2.5)
} else {
  print(p.bundle.all)
}
```

#### 3.2 CN seed overlap across TRACE_cor thresholds

CN seeds that persist only at low ρ are less confident; seeds shared across high ρ values form a stable core set.

```{r}
{
  cn.hit.all <- lapply(c("Negative", "Positive"), function(pol) {
    cn.finder <- data.table::as.data.table(
      trace.demo@advancedAna$TRACE_temp[[pol]]$CNfinder
    )
    if (is.null(cn.finder) || !nrow(cn.finder)) {
      return(NULL)
    }

    lapply(seq_along(cor.list), function(i) {
      cn.hit <- cn.finder[TRACE_cor >= cor.list[i] & is.finite(TRACE_cor)]
      cn.hit[, .SD[which.max(TRACE_cor)], by = from] %>%
        as.data.frame() %>%
        dplyr::mutate(
          cor.cut = cor.list[i],
          param.label = cor.labels[i],
          polarity = pol,
          TRACE_formula = dplyr::coalesce(
            as.character(TRACE_pattern),
            "unknown"
          )
        )
    }) %>% dplyr::bind_rows()
  }) %>%
    dplyr::bind_rows()

  cn.seed <- cn.hit.all %>%
    dplyr::distinct(cor.cut, param.label, polarity, from, .keep_all = TRUE)

  formula.comp <- cn.seed %>%
    dplyr::mutate(
      TRACE_formula = dplyr::coalesce(as.character(TRACE_formula), "unknown"),
      cor.cut = factor(cor.cut, levels = cor.list),
      param.label = factor(param.label, levels = cor.labels)
    ) %>%
    dplyr::count(polarity, cor.cut, param.label, TRACE_formula, name = "feature_n") %>%
    dplyr::ungroup()

  p.formula <- ggplot2::ggplot(
    formula.comp %>% dplyr::filter(polarity == "Negative"),
    ggplot2::aes(x = param.label, y = feature_n, fill = TRACE_formula)
  ) +
    ggplot2::geom_col(position = "stack", width = 0.7) +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
    ggplot2::scale_fill_brewer(palette = "Set2", name = "CN formula") +
    ggplot2::labs(
      x = NULL,
      y = "CN seed count",
      title = "CN seed composition (Negative) across TRACE_cor cutoffs"
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(size = 4, angle = 45, hjust = 1),
      legend.position = "right"
    )

  seed.sets <- lapply(seq_along(cor.list), function(i) {
    cn.seed %>%
      dplyr::filter(cor.cut == cor.list[i]) %>%
      dplyr::pull(from) %>%
      unique()
  })
  names(seed.sets) <- cor.labels

  core.seeds <- Reduce(intersect, seed.sets)
  message(
    "CN seeds shared across all rho = ", length(core.seeds), " / ",
    length(seed.sets[[1]]), " at rho = ", cor.list[1]
  )
}

if (interactive()) {
  open_plot_win(p.formula, 4, 2.5)
} else {
  print(p.formula)
}
```

```{r}
{
  if (requireNamespace("UpSetR", quietly = TRUE)) {
    seed.mat <- UpSetR::fromList(seed.sets)
    p.upset <- UpSetR::upset(
      seed.mat,
      nsets = length(seed.sets),
      order.by = "freq",
      decreasing = TRUE,
      mainbar.y.label = "CN seed intersection size",
      sets.x.label = "CN seeds per TRACE_cor cutoff",
      text.scale = c(1.3, 1.1, 1, 1, 1.2, 1),
      main.bar.color = "#66C2A5",
      sets.bar.color = "#FC8D62"
    )
    print(p.upset)
  } else {
    message(
      "Install UpSetR to draw the CN seed UpSet plot: ",
      "install.packages('UpSetR')"
    )
  }
}
```
