---
title: "noise_evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{noise_evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TRACE)
library(MSdev)
library(dplyr)
library(ggplot2)
```


### 1. Scitific problem

The observation that 'the proportion of high-confidence annotations fell from 2.94% to 1.48%' only reflects true positive results and does not explain the sources of increased “non-biological signals”. Please: (1) describe the blank subtraction method in detail; (2) compare whether blank sample chemical signals change significantly; and (3) discuss other factors contributing to the significant increase in “non-biological signals”. In fact from Supplementary Table 3, when the instrument performance becomes better, annotated metabolites were increased. I suggest modifying the sentence on “2.94% to 1.48%”.


### 2. data set

Example LC-MS demo data (same as `TRACE_Demo`):

```{r}
trace.demo <- MSdev_load("d:/data/PAVE_demo/MSdev_2026_06_01.Rdata")
trace.demo@sampleInfo
```

`trace.demo` is used read-only below. Do not call `MSdev_save(trace.demo)` so the on-disk demo file is not overwritten.

### 3. evaluate

#### 3.1 Blank subtraction (TRACE dynamic filter)

TRACE separates CN-labeled feature pairs from random background pairs in the m/z error (ppm) and RT-shift (s) space of the CN network. For each dimension, `distinct_norm_from_random_backgroud()` fits a Gaussian mixture between CN hits (`cn.hit = TRUE`) and background edges (`cn.hit = FALSE`), then sets a dynamic cutoff at `sd * qnorm(p)` (ppm: 0.999; RT: 0.99999). Pairs exceeding either cutoff are removed before network assignment. This is the blank-subtraction step: background-like CN edges are filtered without subtracting raw blank intensities.

```{r}
{
  i.pol <- 0
  pol <- ifelse(i.pol == 0, "Negative", "Positive")
  temp <- trace.demo@advancedAna$TRACE_temp[[pol]]
  cn.net.eval <- temp$cn.net.eval
  ppm.dyn <- temp$mz.dyn$sd * stats::qnorm(0.999)
  rt.tol.dyn <- temp$rt.dyn$sd * stats::qnorm(0.99999)

  set.seed(1)
  cn.plot.df <- cn.net.eval[
    sample.int(nrow(cn.net.eval), min(20000L, nrow(cn.net.eval))),
  ]
  cols <- c("TRUE" = "red", "FALSE" = "#888888")

  p.dyn <- ggplot2::ggplot(
    cn.plot.df,
    ggplot2::aes(x = mz.ppm, y = rt.diff, col = cn.hit)
  ) +
    ggplot2::geom_point(pch = 16, alpha = 0.08, size = 0.15) +
    ggplot2::geom_vline(
      xintercept = ppm.dyn * c(-1, 1),
      col = "red",
      lty = 2,
      linewidth = 0.4
    ) +
    ggplot2::geom_hline(
      yintercept = rt.tol.dyn * c(-1, 1),
      col = "red",
      lty = 2,
      linewidth = 0.4
    ) +
    ggplot2::scale_color_manual(
      values = cols,
      labels = c("Background edge", "CN-labeled edge"),
      name = NULL
    ) +
    ggplot2::labs(
      x = "m/z error (ppm)",
      y = "RT shift (s)",
      title = paste0("Dynamic filter — ", pol)
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(legend.position = c(0.2, 0.85))

  cn.hit.frac <- mean(cn.net.eval$cn.hit)
  message(
    pol, ": dynamic m/z cutoff = ", round(ppm.dyn, 3), " ppm; ",
    "RT cutoff = ", round(rt.tol.dyn, 2), " s; ",
    "CN-labeled edges = ", sprintf("%.1f%%", 100 * cn.hit.frac)
  )
}

if (interactive()) {
  open_plot_win(p.dyn, 3, 2.5)
} else {
  print(p.dyn)
}
```

#### 3.2 Blank sample chemical signal comparison

Compare mean feature intensity in blank replicates versus biological replicates (four TRACE labeling groups). Features with higher blank than biological signal are potential carry-over or solvent background.

```{r}
{
  i.pol <- 0
  pol <- ifelse(i.pol == 0, "Negative", "Positive")
  xcms.xcms <- trace.demo@xcmsData[[paste0(pol, "MS1")]]
  xcms.val <- xcms::featureValues(xcms.xcms, missing = 0, value = "maxo")
  pda <- Biobase::pData(xcms.xcms)

  blank.samps <- rownames(pda)[pda$group == "Blank"]
  bio.samps <- rownames(pda)[
    pda$group %in% c("S12C14N", "S12C15N", "S13C14N", "S13C15N")
  ]

  feat.df <- data.frame(
    feature_id = seq_len(nrow(xcms.val)),
    blank_mean = rowMeans(xcms.val[, blank.samps, drop = FALSE]),
    bio_mean = rowMeans(xcms.val[, bio.samps, drop = FALSE]),
    blank_rsd = apply(
      xcms.val[, blank.samps, drop = FALSE],
      1,
      function(x) {
        x <- x[is.finite(x) & x > 0]
        if (length(x) < 2) {
          return(NA_real_)
        }
        stats::sd(x) / mean(x)
      }
    )
  ) %>%
    dplyr::mutate(
      log2_fc = log2((bio_mean + 1) / (blank_mean + 1)),
      blank_enriched = blank_mean > bio_mean
    )

  blank.test <- stats::wilcox.test(
    feat.df$blank_mean,
    feat.df$bio_mean,
    paired = TRUE
  )

  p.blank.fc <- ggplot2::ggplot(feat.df, ggplot2::aes(x = log2_fc)) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density), fill = blank_enriched),
      bins = 60,
      alpha = 0.75,
      position = "identity"
    ) +
    ggplot2::geom_vline(xintercept = 0, lty = 2, linewidth = 0.4) +
    ggplot2::scale_fill_manual(
      values = c("TRUE" = "#E41A1C", "FALSE" = "#377EB8"),
      labels = c("Biological > blank", "Blank > biological"),
      name = NULL
    ) +
    ggplot2::labs(
      x = expression(log[2] ~ "(bio mean / blank mean)"),
      y = "Density",
      title = paste0("Blank vs biological intensity — ", pol)
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(legend.position = c(0.75, 0.85))

  p.blank.rsd <- ggplot2::ggplot(
    feat.df %>% dplyr::filter(is.finite(blank_rsd)),
    ggplot2::aes(x = blank_rsd)
  ) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density)),
      bins = 40,
      fill = "#CCCCCC",
      col = "white"
    ) +
    ggplot2::scale_x_continuous(limits = c(0, 2)) +
    ggplot2::labs(
      x = "RSD in blank replicates",
      y = "Density",
      title = "Blank replicate variability"
    ) +
    ggplot2::theme_bw(base_size = 6)

  p.blank <- p.blank.fc / p.blank.rsd +
    patchwork::plot_annotation(tag_levels = "A")

  message(
    pol, ": ", sum(feat.df$blank_enriched), " / ", nrow(feat.df),
    " features have blank mean > biological mean; ",
    "paired Wilcoxon p = ", format.pval(blank.test$p.value, digits = 3)
  )
}

if (interactive()) {
  open_plot_win(p.blank, 4, 2.5)
} else {
  print(p.blank)
}
```

#### 3.3 Annotation composition and noise proportion

Features without a TRACE seed assignment are treated as noise. The stacked bar shows how detected features partition into annotation types versus unassigned noise. The high-confidence metabolite fraction (metabolite seeds / all features) contextualises statements such as “2.94% to 1.48%” — instrument sensitivity increases total features while the noise denominator grows faster than confident metabolite calls.

```{r}
{
  type.levels <- c("metabolite", "adduct", "fragment", "isotope", "noise")
  type.comp.list <- lapply(c("Negative", "Positive"), function(pol) {
    xcms.xcms <- trace.demo@xcmsData[[paste0(pol, "MS1")]]
    n.all <- nrow(xcms::featureDefinitions(xcms.xcms))
    trace.df <- trace.demo@advancedAna$TRACE[[pol]]

    if (is.null(trace.df) || !nrow(trace.df)) {
      return(NULL)
    }

    trace.df <- as.data.frame(trace.df)
    anno.count <- trace.df %>%
      dplyr::mutate(
        type = dplyr::case_when(
          is.na(type) | type == "" ~ "noise",
          TRUE ~ as.character(type)
        )
      ) %>%
      dplyr::count(type, name = "feature_n")

    anno.count <- rbind(
      anno.count,
      data.frame(
        type = "noise",
        feature_n = n.all - sum(anno.count$feature_n),
        stringsAsFactors = FALSE
      )
    ) %>%
      dplyr::group_by(type) %>%
      dplyr::summarise(feature_n = sum(feature_n), .groups = "drop") %>%
      dplyr::mutate(
        polarity = pol,
        type = factor(type, levels = type.levels),
        pct = 100 * feature_n / n.all
      )

    cn.hit <- trace.demo@advancedAna$TRACE_temp[[pol]]$cn.net.hit
    n.high.cor <- sum(cn.hit$TRACE_cor >= 0.75, na.rm = TRUE)

    message(
      pol, ": ", n.all, " features; ",
      "metabolite = ", anno.count$feature_n[anno.count$type == "metabolite"], " (",
      round(anno.count$pct[anno.count$type == "metabolite"], 2), "%); ",
      "noise = ", anno.count$feature_n[anno.count$type == "noise"], " (",
      round(anno.count$pct[anno.count$type == "noise"], 2), "%); ",
      "CN seeds with TRACE_cor >= 0.75 = ", n.high.cor, " (",
      round(100 * n.high.cor / n.all, 2), "%)"
    )

    anno.count
  })

  type.comp <- dplyr::bind_rows(type.comp.list)

  p.type <- ggplot2::ggplot(type.comp, ggplot2::aes(x = polarity, y = feature_n, fill = type)) +
    ggplot2::geom_col(position = "stack", width = 0.65) +
    ggplot2::geom_text(
      ggplot2::aes(label = ifelse(pct >= 3, paste0(round(pct, 1), "%"), "")),
      position = ggplot2::position_stack(vjust = 0.5),
      size = 2,
      col = "white"
    ) +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
    ggplot2::scale_fill_brewer(palette = "Set2", name = "Annotation") +
    ggplot2::labs(
      x = NULL,
      y = "Feature count",
      title = "TRACE annotation vs noise"
    ) +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::theme(legend.position = "right")
}

if (interactive()) {
  open_plot_win(p.type, 3, 2.5)
} else {
  print(p.type)
}
```

#### 3.4 Feature RSD across sample groups (including Blank)

Replicate RSD distinguishes stable biological CN patterns from unstable blank or solvent noise. Blank and QC groups are coloured as noise in the TRACE colour scheme.

```{r}
{
  p.rsd <- plot_TRACE_RSD(
    trace.demo,
    i.pol = 0,
    exclude.blank = FALSE,
    rsd.max = 1.5
  )
}

if (interactive()) {
  open_plot_win(p.rsd, 3, 2.5)
} else {
  print(p.rsd)
}
```

#### 3.5 Other factors contributing to non-biological signals

Besides blank subtraction and raw blank intensity, TRACE noise can arise from: (1) xcms peak picking sensitivity (more features at higher instrument performance), (2) CN edges failing the dynamic m/z–RT filter, (3) low `TRACE_cor` CN bundles, and (4) network re-assignment of adduct/fragment/isotope candidates. Labeling-fraction distributions of confident metabolites help separate true CN labeling from random co-elution.

```{r}
{
  p.label <- plot_TRACE_labeling_fraction(trace.demo, i.pol = 0)
}

if (interactive()) {
  open_plot_win(p.label, 4, 2.5)
} else {
  print(p.label)
}
```

