---
title: "TRACE_Demo"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
vignette: "%\\VignetteIndexEntry{TRACE_Demo} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Data set preparation

### 1.1 load data set

Here is a data set with 15 samples grouped as:

- $^{12}\text{C}$ and $^{14}\text{N}$
- $^{12}\text{C}$ and $^{14}\text{N}$
- $^{13}\text{C}$ and $^{15}\text{N}$
- $^{12}\text{C}$ and $^{15}\text{N}$
- Blank

Use `MSdev` to load the raw data and convert it to `mzML` files.

```{r}
library(MSdev)
demo.dir <- "d:/data/PAVE_demo/rawdata"
trace.demo <- MSdev( rawDataDir =  demo.dir)
trace.demo <- MSdev_msConvert(trace.demo)
trace.demo@sampleInfo
```

### 1.2 Peaks extraction by xcms

`MSdev_xcmsProcessing` call `xcms::findChromPeaks` and `xcms::groupChromPeaks` to extract peaks and group peaks.

Use `MSdev_checkSampleInfo` to input the sample information: sample.group

Use `MSdev_set_param` to set the parameters for `xcms::findChromPeaks` and `xcms::groupChromPeaks` and then run `MSdev_xcmsProcessing` to extract peaks and group peaks.

```{r}
if(F){
  trace.demo <- MSdev_set_param(
    trace.demo,
    findChromPeaks =
      xcms::CentWaveParam(
        ppm = 5,
        prefilter = c(3,1000),
        peakwidth = c(10,30),
        snthresh = 10,
        fitgauss = T),
    groupChromPeaks =
      xcms::PeakDensityParam(
        sampleGroups = "A",
        minFraction = 0.6,
        binSize = 0.002,
        bw = 5,
        ppm = 5))
  trace.demo <- MSdev_checkSampleInfo(trace.demo)
  trace.demo <- MSdev_xcmsProcessing(trace.demo)
  
}
```

Save the processed data set to `Rdata` file, so that you can load the processed data set later without running the `MSdev_xcmsProcessing` again. Load the processed data set from `Rdata` file:

```{r}
#MSdev_save(trace.demo)

trace.demo <- MSdev_load("d:/data/PAVE_demo/MSdev_2026_06_01.Rdata")
trace.demo
```

## 2. TRACE Core

### 2.1 find CN labelling

```{r}
trace.demo <- TRACE_get_CN_net(trace.demo,0,ratio.adjust = c(1.0842,0.7641,1.0566,1.0707))
trace.demo <- TRACE_dynamic_filter(trace.demo,0)
trace.demo <- TRACE_network_assignment(trace.demo,0)
trace.demo <- TRACE_annotate(trace.demo,i.pol = 0)

trace.demo <- TRACE_get_CN_net(trace.demo,1,ratio.adjust = c(1.0842,0.7641,1.0566,1.0707))
trace.demo <- TRACE_dynamic_filter(trace.demo,1)
trace.demo <- TRACE_network_assignment(trace.demo,1)
trace.demo <- TRACE_annotate(trace.demo,i.pol = 1)

```

### 2.x
