---
title: "TRACE Nutrient Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TRACE Nutrient Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette compares **absorb-form preference** across eight
nutrient-supplement conditions. For each TRACE metabolite (e.g.
`C10N5`), all CN labeling forms (`C0N0`, `C7N3`, …) are matched by **m/z
and RT**, and each labeling form is converted to its **absorb form** —
the unlabeled C/N atoms incorporated from nutrients:

$$\text{absorb form } C_{x'}N_{y'} = C_{(C_{\max} - x)}N_{(N_{\max} - y)}$$

For example, metabolite `C10N5` with labeling form `C7N3` has absorb
form `C3N2` (3 carbons and 2 nitrogens absorbed from unlabeled sources).

The workflow follows the nutrient experiment analysis in `test/test.R`
and extends it from aggregate C/N absorption fractions to
**per–labeling-form absorb pattern** distributions across nutrient
conditions.

```{r setup}
library(TRACE)
library(MSdev)
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(tibble)
library(SummarizedExperiment)
```

## 1. Load data

```{r load-data}
nutrient.file <- "d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata"

if (!file.exists(nutrient.file)) {
  stop(
    "Nutrition MSdev object not found at:\n  ", nutrient.file, "\n",
    "Update `nutrient.file` to your local path."
  )
}

obj <- MSdev_load(nutrient.file)

nutrient.groups <- c("NSA", "NSB", "NSC", "NSD", "NSE", "NSF", "NSG", "NSH")
nutrient.labels <- c(
  NSA = "Control",
  NSB = "+mixture",
  NSC = "+leucine",
  NSD = "+threonine",
  NSE = "+tryptophan",
  NSF = "+adenine",
  NSG = "+uracil",
  NSH = "+acetate"
)
nutrient.colors <- c(
  "#E64B35", "#4DBBD5", "#00A087", "#3C5488",
  "#F39B7F", "#8491B4", "#91D1C2", "#7E6148"
)
names(nutrient.colors) <- nutrient.groups
```

## 2. Helper functions

### Match m/z / RT and extract all labeling forms

```{r helpers-extract}
extract_cn_labeling_forms <- function(
    pave_formula,
    seed_mz,
    seed_rt,
    xcms_fdf,
    xcms_mat,
    xcms_se,
    ppm = 3,
    rt_tol = 5) {
  cn_count <- MSCC::chemform_parse(pave_formula)
  c_max <- cn_count[1, "C"]
  n_max <- cn_count[1, "N"]

  cn_diff <- TRACE:::get_CN_mass_diff_table(c_max, n_max)
  cn_diff[, mz := mass_diff + seed_mz]

  cn_match <- MSdev::match_mz_foverlaps(
    cn_diff$mz,
    xcms_fdf$mzmed,
    ppm = ppm
  )
  cn_match[, fid := ion2][, rt := xcms_fdf$rtmed[fid]]

  cn_table <- cbind(cn_match, cn_diff[cn_match$ion1])
  cn_table[, label_pattern := paste0("C", C_count, "N", N_count)][!is.na(fid)]

  cn_table <- cn_table[abs(rt - seed_rt) < rt_tol]
  cn_table <- cn_table %>%
    dplyr::group_by(label_pattern) %>%
    dplyr::slice_min(mz.ppm, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(N_count, C_count)

  if (nrow(cn_table) <= 1) {
    return(NULL)
  }

  exp_mat <- xcms_mat[cn_table$fid, xcms_se$sample.name, drop = FALSE]
  rownames(exp_mat) <- cn_table$label_pattern

  list(
    cn_table = as.data.frame(cn_table),
    exp_mat = exp_mat,
    c_max = c_max,
    n_max = n_max
  )
}
```

### Calculate absorb form fractions per sample

For each labeling form `CxNy`, the absorb form is
`C(Cmax − x)N(Nmax − y)`. Intensities are normalized to the control
group (`NSA`) and converted to fractions within each sample.

```{r helpers-absorb}
compute_absorb_forms <- function(
    cn_result,
    xcms_se,
    norm_group = "NSA",
    norm_replicates = 10) {
  cn_table <- cn_result$cn_table
  exp_mat <- cn_result$exp_mat
  c_max <- cn_result$c_max
  n_max <- cn_result$n_max

  norm_samples <- xcms_se$sample.name[xcms_se$group == norm_group]
  norm_to <- sum(exp_mat[, norm_samples, drop = FALSE], na.rm = TRUE) / norm_replicates
  if (!is.finite(norm_to) || norm_to <= 0) {
    norm_to <- mean(exp_mat, na.rm = TRUE)
  }
  exp_mat <- exp_mat / norm_to

  cn_table$c_absorb <- c_max - cn_table$C_count
  cn_table$n_absorb <- n_max - cn_table$N_count
  cn_table$absorb_pattern <- paste0("C", cn_table$c_absorb, "N", cn_table$n_absorb)

  exp_mat %>%
    as.data.frame() %>%
    tibble::rownames_to_column("label_pattern") %>%
    pivot_longer(-label_pattern, names_to = "sample", values_to = "intensity") %>%
    left_join(cn_table, by = "label_pattern") %>%
    group_by(sample) %>%
    mutate(
      group = unname(setNames(xcms_se$sample.type, xcms_se$sample.name)[sample]),
      int_total = sum(intensity, na.rm = TRUE),
      absorb_fraction = ifelse(int_total > 0, intensity / int_total, NA_real_)
    ) %>%
    ungroup() %>%
    mutate(
      c_max = c_max,
      n_max = n_max
    )
}
```

Fully labeled metabolites have absorb form `C0N0` (no unlabeled C/N
incorporated). Exclude this pattern when visualizing absorb-form
preference and renormalize remaining fractions to sum to 1 within each
metabolite sample.

```{r helpers-filter}
drop_absorb_c0n0 <- function(df, renormalize = TRUE) {
  df <- df %>%
    dplyr::filter(absorb_pattern != "C0N0")

  if (!renormalize || !nrow(df)) {
    return(df)
  }

  renorm_keys <- intersect(c("feature_id", "polarity", "sample"), names(df))
  if (!length(renorm_keys)) {
    renorm_keys <- intersect(c("feature_id", "polarity", "group"), names(df))
  }

  df %>%
    group_by(across(all_of(renorm_keys))) %>%
    mutate(
      absorb_fraction = absorb_fraction / sum(absorb_fraction, na.rm = TRUE)
    ) %>%
    ungroup()
}
```

### Run absorb-form analysis for one polarity

```{r helpers-run}
run_absorb_form_analysis <- function(
    obj,
    i.pol = 1,
    ppm = 3,
    rt_tol = 5,
    min_peak_maxo = 1e4,
    nutrient_groups = nutrient.groups) {
  pol <- if (i.pol == 0) "Negative" else "Positive"
  xcms_obj <- obj@xcmsData[[paste0(pol, "MS1")]]
  pave_seed <- obj@statData$TRACE[[pol]]
  if (!is.data.frame(pave_seed)) {
    pave_seed <- as.data.frame(pave_seed)
  }
  pave_seed <- pave_seed %>%
    dplyr::filter(type == "metabolite")

  xcms_se_raw <- MSdev::get_xcms_feature_se(xcms_obj, missing = 1)
  xcms_se <- xcms_se_raw[, grepl("NS", xcms_se_raw$sample.type)]
  xcms_se <- MSdev::se_adjuset_by_weight(xcms_se)

  xcms_fdf <- as.data.frame(rowData(xcms_se))
  xcms_fdf$peakMaxo <- rowMeans(
    assay(xcms_se_raw[, xcms_se_raw$group == "S12C14N", drop = FALSE])
  )
  xcms_mat <- assay(xcms_se)

  results <- vector("list", nrow(pave_seed))
  pb <- MSdev:::get_progress_bar(nrow(pave_seed))

  for (i in seq_len(nrow(pave_seed))) {
    pb$tick()
    i_fid <- as.numeric(pave_seed$feature_id[i])
    if (!is.finite(i_fid) || i_fid < 1 || i_fid > nrow(xcms_fdf)) next

    cn_result <- extract_cn_labeling_forms(
      pave_formula = pave_seed$pave_formula[i],
      seed_mz = xcms_fdf$mzmed[i_fid],
      seed_rt = pave_seed$rt[i],
      xcms_fdf = xcms_fdf,
      xcms_mat = xcms_mat,
      xcms_se = xcms_se,
      ppm = ppm,
      rt_tol = rt_tol
    )
    if (is.null(cn_result)) next

    absorb_df <- compute_absorb_forms(cn_result, xcms_se)
    absorb_df$feature_id <- pave_seed$feature_id[i]
    absorb_df$compound_id <- if ("compound_id" %in% names(pave_seed)) {
      pave_seed$compound_id[i]
    } else {
      NA_character_
    }
    absorb_df$fid <- paste0(tolower(substr(pol, 1, 3)), i_fid)
    absorb_df$metabolite <- pave_seed$name[i]
    absorb_df$kegg_id <- if ("kegg_id" %in% names(pave_seed)) pave_seed$kegg_id[i] else NA_character_
    absorb_df$pave_formula <- pave_seed$pave_formula[i]
    absorb_df$peak_maxo <- xcms_fdf$peakMaxo[i_fid]
    absorb_df$polarity <- pol
    results[[i]] <- absorb_df
  }

  bind_rows(results) %>%
    filter(
      group %in% nutrient_groups,
      peak_maxo > min_peak_maxo,
      is.finite(absorb_fraction)
    )
}
```

## 3. Extract absorb forms for all metabolites

```{r run-analysis, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
absorb.pos <- run_absorb_form_analysis(obj, i.pol = 1)
absorb.neg <- run_absorb_form_analysis(obj, i.pol = 0)
absorb.all <- bind_rows(absorb.pos, absorb.neg)
absorb.all.plot <- drop_absorb_c0n0(absorb.all)

absorb.summary <- absorb.all.plot %>%
  group_by(
    feature_id, compound_id, metabolite, pave_formula, kegg_id, polarity,
    group, absorb_pattern, c_absorb, n_absorb
  ) %>%
  summarize(
    absorb_fraction = mean(absorb_fraction, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels)
  )

head(absorb.summary)
```

## 4. Plot absorb-form preference

### 4.1 Example metabolite: ATP (`C10N5`)

Labeling form `C7N3` → absorb form `C3N2`. The heatmap shows the
relative abundance of each absorb pattern in the eight nutrient
conditions (`C0N0` excluded — fully labeled, no nutrient absorption).

```{r plot-example, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
example_pick <- absorb.summary %>%
  filter(
    (!is.na(metabolite) & grepl("NAD|C10N5", metabolite, ignore.case = TRUE))
  ) %>%
  distinct(feature_id, metabolite, pave_formula) %>%
  dplyr::slice_head(n = 1)

if (nrow(example_pick) == 0) {
  example_pick <- absorb.summary %>%
    group_by(feature_id, metabolite, pave_formula) %>%
    summarize(n_absorb = n_distinct(absorb_pattern), .groups = "drop") %>%
    filter(n_absorb >= 4) %>%
    dplyr::slice_head(n = 1)
}

example_met <- absorb.summary %>%
  filter(
    feature_id == example_pick$feature_id[1],
    pave_formula == example_pick$pave_formula[1]
  )

example_met <- example_met %>%
  mutate(
    absorb_pattern = factor(
      absorb_pattern,
      levels = unique(absorb_pattern[order(c_absorb, n_absorb)])
    )
  )

ggplot(example_met, aes(x = nutrient, y = absorb_pattern, fill = absorb_fraction)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "#FFFFFF",
    mid = "#FFCD99",
    high = "#D13B2E",
    midpoint = max(example_met$absorb_fraction, na.rm = TRUE) / 2,
    name = "Fraction"
  ) +
  labs(
    title = paste0(example_met$metabolite[1], " (", example_met$pave_formula[1], ")"),
    subtitle = "Absorb-form preference across nutrient conditions",
    x = NULL,
    y = "Absorb form"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = -35, hjust = 0),
    panel.grid = element_blank()
  )
```

### 4.2 Stacked bar: absorb-form composition per condition

For the same metabolite, stacked bars show how the absorb-form profile
shifts across nutrient supplements.

```{r plot-stacked, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
if (exists("example_met") && nrow(example_met) > 0) {
  ggplot(example_met, aes(x = nutrient, y = absorb_fraction, fill = absorb_pattern)) +
    geom_col(color = "grey30", linewidth = 0.2) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    labs(
      title = paste0(example_met$metabolite[1], " absorb-form composition"),
      x = NULL,
      y = "Fraction of total CN signal",
      fill = "Absorb form"
    ) +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = -35, hjust = 0))
}
```

### 4.3 Global comparison across nutrient conditions

#### 4.3.1 C/N absorption

**True absorption fraction** (as in `test/test.R`): weighted mean
absorbed C or N atoms per metabolite, including the fully labeled absorb
form `C0N0`. Higher values mean more unlabeled nutrient incorporation.
The scatter plot summarizes mean absorbed C and N counts per metabolite
sample (`C0N0` excluded).

```{r plot-global-cn-absorption, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
true.absorb <- absorb.all %>%
  filter(group %in% nutrient.groups, c_max > 0) %>%
  group_by(feature_id, polarity, sample, group) %>%
  summarize(
    c_absorb_frac = sum(c_absorb * absorb_fraction, na.rm = TRUE) / dplyr::first(c_max),
    n_absorb_frac = if (dplyr::first(n_max) > 0) {
      sum(n_absorb * absorb_fraction, na.rm = TRUE) / dplyr::first(n_max)
    } else {
      NA_real_
    },
    c0n0_frac = sum(absorb_fraction[absorb_pattern == "C0N0"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels)
  )

p_true_c <- ggplot(true.absorb, aes(x = nutrient, y = c_absorb_frac, fill = group)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_fill_manual(values = nutrient.colors, guide = "none") +
  labs(
    title = "True C absorption fraction (all absorb forms)",
    subtitle = "Mixture (+mixture) typically exceeds control when supplements are used",
    x = NULL,
    y = "Absorption fraction of C"
  ) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = -35, hjust = 0))

p_true_n <- ggplot(
  true.absorb %>% filter(is.finite(n_absorb_frac)),
  aes(x = nutrient, y = n_absorb_frac, fill = group)
) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_fill_manual(values = nutrient.colors, guide = "none") +
  labs(
    title = "True N absorption fraction (all absorb forms)",
    x = NULL,
    y = "Absorption fraction of N"
  ) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = -35, hjust = 0))

global.absorb <- absorb.all.plot %>%
  filter(group %in% nutrient.groups) %>%
  group_by(feature_id, compound_id, metabolite, pave_formula, kegg_id, sample, group) %>%
  summarize(
    c_absorb_mean = sum(c_absorb * absorb_fraction, na.rm = TRUE),
    n_absorb_mean = sum(n_absorb * absorb_fraction, na.rm = TRUE),
    int_total = dplyr::first(int_total),
    .groups = "drop"
  ) %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels)
  )

p_scatter <- ggplot(global.absorb, aes(x = c_absorb_mean, y = n_absorb_mean, color = group, fill = group)) +
  geom_point(size = 0.8, alpha = 0.35) +
  stat_ellipse(geom = "polygon", alpha = 0.15, linewidth = 0.4) +
  scale_color_manual(values = nutrient.colors, labels = nutrient.labels) +
  scale_fill_manual(values = nutrient.colors, labels = nutrient.labels) +
  labs(
    title = "Absorb-form preference across metabolites",
    subtitle = "Mean absorbed C and N counts (C0N0 excluded)",
    x = "Mean absorbed C count",
    y = "Mean absorbed N count",
    color = "Treatment",
    fill = "Treatment"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "right")

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_true_c, p_true_n, p_scatter, ncol = 1, heights = c(1, 1, 1.2))
} else {
  print(p_true_c)
  print(p_true_n)
  print(p_scatter)
}
```

#### 4.3.2 CN form absorption

**Absorb-form composition** (after dropping `C0N0`): relative abundance
of partially labeled absorb patterns only. Sum absorb ratio of the **top
20 absorb forms** plus **Other** in each nutrient condition (stacked bar
heights are equal per condition). The top-20 set is ranked globally;
mixture often shows a **lower top-20 slice** because more absorb signal
is distributed outside those 20 patterns, not because total absorption
is lower (see §4.3.1).

The heatmap uses **ComplexHeatmap** with circle markers in each cell.
Values are row-normalized so each absorb form shows its **relative
absorption** across the eight conditions (proportions sum to 1 per row).

```{r plot-global-cn-form, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
absorb.global <- absorb.all.plot %>%
  filter(group %in% nutrient.groups) %>%
  group_by(feature_id, polarity, group, absorb_pattern, c_absorb, n_absorb) %>%
  summarize(
    mean_absorb_ratio = mean(absorb_fraction, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(group, absorb_pattern, c_absorb, n_absorb) %>%
  summarize(
    sum_absorb_ratio = sum(mean_absorb_ratio, na.rm = TRUE),
    .groups = "drop"
  )

top20.absorb <- absorb.global %>%
  group_by(absorb_pattern) %>%
  summarize(total = sum(sum_absorb_ratio, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  dplyr::slice_head(n = 20)

absorb.top20 <- absorb.global %>%
  mutate(
    absorb_display = if_else(
      absorb_pattern %in% top20.absorb$absorb_pattern,
      absorb_pattern,
      "Other"
    )
  ) %>%
  group_by(group, absorb_display) %>%
  summarize(sum_absorb_ratio = sum(sum_absorb_ratio, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels),
    absorb_display = if_else(
      absorb_display == "Other",
      "Other",
      absorb_display
    ),
    absorb_display = factor(
      absorb_display,
      levels = c(top20.absorb$absorb_pattern, "Other")
    )
  )

ggplot(absorb.top20, aes(x = nutrient, y = sum_absorb_ratio, fill = absorb_display)) +
  geom_col(color = "grey30", linewidth = 0.15, width = 0.75) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03))) +
  scale_fill_manual(
    values = c(
      setNames(
        viridisLite::viridis(length(top20.absorb$absorb_pattern), option = "turbo", direction = -1),
        top20.absorb$absorb_pattern
      ),
      Other = "#CCCCCC"
    )
  ) +
  labs(
    title = "Top 20 absorb forms + Other across nutrient conditions",
    subtitle = "Equal bar height per condition; lower top-20 slice in mixture reflects broader absorb-form usage",
    x = NULL,
    y = "Sum absorb ratio",
    fill = "Absorb form"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = -35, hjust = 0),
    legend.key.size = unit(0.35, "cm"),
    legend.text = element_text(size = 8)
  )

absorb.top20.detail <- absorb.global %>%
  filter(absorb_pattern %in% top20.absorb$absorb_pattern) %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels),
    absorb_pattern = factor(absorb_pattern, levels = top20.absorb$absorb_pattern)
  )

absorb.top20.order <- absorb.global %>%
  filter(absorb_pattern %in% top20.absorb$absorb_pattern) %>%
  distinct(absorb_pattern, c_absorb, n_absorb) %>%
  arrange(c_absorb, n_absorb)

absorb.top20.hm <- absorb.top20.detail %>%
  left_join(absorb.top20.order, by = "absorb_pattern") %>%
  group_by(absorb_pattern) %>%
  mutate(
    relative_absorption = sum_absorb_ratio / sum(sum_absorb_ratio, na.rm = TRUE)
  ) %>%
  ungroup()

if (requireNamespace("ComplexHeatmap", quietly = TRUE)) {
  library(ComplexHeatmap)

  hm_mat.top20 <- absorb.top20.hm %>%
    select(absorb_pattern, group, relative_absorption) %>%
    pivot_wider(names_from = group, values_from = relative_absorption) %>%
    tibble::column_to_rownames("absorb_pattern") %>%
    as.matrix()

  hm_mat.top20 <- hm_mat.top20[
    absorb.top20.order$absorb_pattern,
    nutrient.groups,
    drop = FALSE
  ]
  hm_mat.top20[is.na(hm_mat.top20)] <- 0

  col_fun.top20 <- circlize::colorRamp2(
    c(0, 0.5, 1),
    c("#FFFFFF", "#FFCD99", "#D13B2E")
  )

  ht.top20 <- Heatmap(
    hm_mat.top20,
    name = "Relative\nabsorption",
    col = col_fun.top20,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    column_title = "Relative absorption of top 20 absorb forms",
    column_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
    column_labels = nutrient.labels[nutrient.groups],
    column_names_rot = 30,
    row_names_gp = grid::gpar(fontsize = 9),
    column_names_gp = grid::gpar(fontsize = 9),
    rect_gp = grid::gpar(type = "none"),
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid.circle(
        x = x,
        y = y,
        r = min(unit.c(width, height)) * 0.42,
        gp = grid::gpar(fill = fill, col = fill)
      )
    },
    heatmap_legend_param = list(
      title_gp = grid::gpar(fontsize = 9),
      labels_gp = grid::gpar(fontsize = 8)
    )
  )

  draw(ht.top20)
} else {
  message("Install ComplexHeatmap for the top-20 absorb-form dot heatmap.")
}
```

### 4.4 Boxplot: absorbed C fraction by nutrient condition

Compare the fraction of unlabeled (absorbed) carbon across all
metabolites, mirroring the boxplot panel in the nutrient comparison of
`test/test.R`.

```{r plot-boxplot, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
met.absorb.frac <- absorb.all.plot %>%
  filter(group %in% nutrient.groups) %>%
  group_by(feature_id, compound_id, metabolite, kegg_id, sample, group) %>%
  summarize(
    c_absorb_frac = sum(c_absorb * absorb_fraction, na.rm = TRUE) / dplyr::first(c_max),
    n_absorb_frac = sum(n_absorb * absorb_fraction, na.rm = TRUE) / dplyr::first(n_max),
    .groups = "drop"
  ) %>%
  filter(c_absorb_frac > 0.02 | n_absorb_frac > 0.02) %>%
  mutate(
    nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels)
  )

p_c <- ggplot(met.absorb.frac, aes(x = nutrient, y = c_absorb_frac, fill = group)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_fill_manual(values = nutrient.colors, guide = "none") +
  labs(x = NULL, y = "Absorption fraction of C", tag = "C") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = -35, hjust = 0))

p_n <- ggplot(met.absorb.frac, aes(x = nutrient, y = n_absorb_frac, fill = group)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_fill_manual(values = nutrient.colors, guide = "none") +
  labs(x = NULL, y = "Absorption fraction of N", tag = "N") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_blank())

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_c, p_n, ncol = 1)
} else {
  print(p_c)
  print(p_n)
}
```

### 4.5 Heatmap panel for selected metabolites

Reproduce the per-metabolite labeling-form heatmaps from `test/test.R`,
with rows annotated by both **labeling form** and **absorb form**.

```{r plot-heatmap-panel, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
selected <- absorb.all.plot %>%
  filter(
    pave_formula %in% c("C10N5", "C5N1", "C6N2", "C10N3", "C21N7", "C5N0") |
      metabolite %in% c("ATP", "Glutamate", "Glutamine", "GSH", "NAD", "R5P")
  ) %>%
  distinct(feature_id, metabolite, pave_formula, polarity) %>%
  dplyr::slice_head(n = 6)

if (nrow(selected) > 0 && requireNamespace("ComplexHeatmap", quietly = TRUE)) {
  library(ComplexHeatmap)

  for (i in seq_len(nrow(selected))) {
    sub <- absorb.all.plot %>%
      filter(
        feature_id == selected$feature_id[i],
        polarity == selected$polarity[i]
      )

    hm_mat <- sub %>%
      group_by(group, label_pattern, absorb_pattern) %>%
      summarize(fraction = mean(absorb_fraction, na.rm = TRUE), .groups = "drop") %>%
      mutate(row_label = paste0(label_pattern, " → ", absorb_pattern)) %>%
      select(row_label, group, fraction) %>%
      pivot_wider(names_from = group, values_from = fraction) %>%
      tibble::column_to_rownames("row_label") %>%
      as.matrix()

    hm_mat <- hm_mat[, nutrient.groups[nutrient.groups %in% colnames(hm_mat)], drop = FALSE]
    hm_mat[is.na(hm_mat)] <- 0

    col_fun <- circlize::colorRamp2(
      c(0, max(hm_mat) / 2, max(hm_mat)),
      c("#FFFFFF", "#FFCD99", "#D13B2E")
    )

    ht <- Heatmap(
      hm_mat,
      name = "Fraction",
      col = col_fun,
      cluster_rows = FALSE,
      cluster_columns = FALSE,
      column_title = paste0(selected$metabolite[i], " (", selected$pave_formula[i], ")"),
      column_title_gp = grid::gpar(fontsize = 9),
      row_names_gp = grid::gpar(fontsize = 7),
      column_labels = nutrient.labels[colnames(hm_mat)]
    )
    draw(ht)
  }
} else if (nrow(selected) == 0) {
  message("No selected metabolites found in absorb results.")
} else {
  message("Install ComplexHeatmap for multi-metabolite heatmap panels.")
}
```

### 4.6 Differential metabolites by top-20 absorb profile

Each metabolite is represented as a **length-20 vector** of mean absorb
ratios for the globally defined top 20 absorb forms (missing forms = 0).
For every metabolite, an **8 × 20** matrix is built (nutrient groups ×
absorb forms). Metabolites are ranked by the maximum pairwise Euclidean
distance between group centroids (`diff_score`); higher scores indicate
stronger absorb-profile shifts across nutrient conditions.

```{r plot-diff-metabolites, eval = file.exists("d:/data/2026.01.07.PAVE.Nutrition/MSdev_2026_01_07.Rdata")}
if (!exists("top20.absorb")) {
  top20.absorb <- absorb.all.plot %>%
    filter(group %in% nutrient.groups) %>%
    group_by(feature_id, polarity, group, absorb_pattern) %>%
    summarize(mean_absorb_ratio = mean(absorb_fraction, na.rm = TRUE), .groups = "drop") %>%
    group_by(absorb_pattern) %>%
    summarize(total = sum(mean_absorb_ratio, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(total)) %>%
    dplyr::slice_head(n = 50)
}

top20.forms <- top20.absorb$absorb_pattern

met_absorb_vector <- absorb.all.plot %>%
  filter(
    group %in% nutrient.groups,
    absorb_pattern %in% top20.forms
  ) %>%
  group_by(
    feature_id, polarity, metabolite, compound_id, pave_formula, kegg_id,
    group, absorb_pattern
  ) %>%
  summarize(absorb_ratio = mean(absorb_fraction, na.rm = TRUE), .groups = "drop") %>%
  tidyr::complete(
    tidyr::nesting(
      feature_id, polarity, metabolite, compound_id, pave_formula, kegg_id
    ),
    group = nutrient.groups,
    absorb_pattern = top20.forms,
    fill = list(absorb_ratio = 0)
  )

met_absorb_wide <- met_absorb_vector %>%
  pivot_wider(
    id_cols = c(
      feature_id, polarity, metabolite, compound_id, pave_formula, kegg_id, group
    ),
    names_from = absorb_pattern,
    values_from = absorb_ratio
  )

met_label <- function(metabolite, pave_formula, feature_id, polarity) {
  nm <- dplyr::coalesce(as.character(metabolite), as.character(pave_formula))
  paste0(nm, " (", polarity, " #", feature_id, ")")
}

met_diff <- met_absorb_wide %>%
  mutate(
    met_label = met_label(metabolite, pave_formula, feature_id, polarity)
  ) %>%
  group_by(feature_id, polarity, metabolite, compound_id, pave_formula, kegg_id, met_label) %>%
  summarize(
    diff_score = {
      mat <- as.matrix(dplyr::across(dplyr::all_of(top20.forms)))
      if (nrow(mat) < 2) 0 else max(stats::dist(mat))
    },
    sum_var = {
      mat <- as.matrix(dplyr::across(dplyr::all_of(top20.forms)))
      sum(apply(mat, 2, stats::var, na.rm = TRUE))
    },
    absorb_profile = list({
      mat <- as.matrix(dplyr::across(dplyr::all_of(top20.forms)))
      stats::setNames(colMeans(mat), top20.forms)
    }),
    .groups = "drop"
  ) %>%
  arrange(desc(diff_score), desc(sum_var))

top_n_diff <- 30
top_diff_mets <- met_diff %>%
  dplyr::slice_head(n = top_n_diff)

knitr::kable(
  top_diff_mets %>%
    select(
      met_label, feature_id, compound_id, pave_formula, polarity,
      diff_score, sum_var
    ) %>%
    dplyr::slice_head(n = 20),
  caption = "Top 20 metabolites ranked by absorb-profile distance across nutrient groups"
)

p_diff_bar <- ggplot(
  top_diff_mets %>% dplyr::slice_head(n = 20),
  aes(x = reorder(met_label, diff_score), y = diff_score, fill = polarity)
) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Most differential metabolites (top-20 absorb profile)",
    x = NULL,
    y = "Max pairwise distance across 8 groups",
    fill = "Polarity"
  ) +
  theme_bw(base_size = 6) +
  theme(legend.position = "bottom")

print(p_diff_bar)

if (requireNamespace("ComplexHeatmap", quietly = TRUE)) {
  hm_diff <- met_absorb_vector %>%
    inner_join(
      top_diff_mets %>%
        select(feature_id, polarity, met_label),
      by = c("feature_id", "polarity")
    ) %>%
    mutate(
      col_key = paste0(group, "|", absorb_pattern),
      met_label = factor(met_label, levels = rev(unique(top_diff_mets$met_label)))
    ) %>%
    select(met_label, col_key, absorb_ratio) %>%
    pivot_wider(names_from = col_key, values_from = absorb_ratio) %>%
    tibble::column_to_rownames("met_label") %>%
    as.matrix()

  col_order <- as.vector(t(outer(nutrient.groups, top20.forms, paste, sep = "|")))
  col_order <- col_order[col_order %in% colnames(hm_diff)]
  hm_diff <- hm_diff[, col_order, drop = FALSE]
  hm_diff[is.na(hm_diff)] <- 0

  col_fun_diff <- circlize::colorRamp2(
    c(0, max(hm_diff) / 2, max(hm_diff)),
    c("#FFFFFF", "#FFCD99", "#D13B2E")
  )

  col_split_diff <- factor(
    gsub("\\|.*$", "", colnames(hm_diff)),
    levels = nutrient.groups
  )

  ht_diff <- Heatmap(
    hm_diff,
    name = "Absorb\nratio",
    col = col_fun_diff,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    column_split = col_split_diff,
    column_title_gp = grid::gpar(fontsize = 8),
    column_labels = gsub("^.*\\|", "", colnames(hm_diff)),
    column_names_rot = 90,
    row_names_gp = grid::gpar(fontsize = 7),
    column_names_gp = grid::gpar(fontsize = 6),
    rect_gp = grid::gpar(type = "none"),
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid.circle(
        x = x,
        y = y,
        r = min(unit.c(width, height)) * 0.42,
        gp = grid::gpar(fill = fill, col = fill)
      )
    },
    column_title = "Top differential metabolites: top-20 absorb ratio profile",
    heatmap_legend_param = list(
      title_gp = grid::gpar(fontsize = 9),
      labels_gp = grid::gpar(fontsize = 8)
    )
  )

  draw(ht_diff)
} else {
  hm_gg <- met_absorb_vector %>%
    inner_join(
      top_diff_mets %>% select(feature_id, polarity, met_label),
      by = c("feature_id", "polarity")
    ) %>%
    mutate(
      nutrient = factor(group, levels = nutrient.groups, labels = nutrient.labels),
      met_label = factor(met_label, levels = rev(unique(top_diff_mets$met_label)))
    )

  ggplot(hm_gg, aes(x = absorb_pattern, y = met_label, fill = absorb_ratio)) +
    geom_tile(color = "white", linewidth = 0.2) +
    facet_wrap(~ nutrient, nrow = 1, scales = "free_x") +
    scale_fill_gradient2(
      low = "#FFFFFF", mid = "#FFCD99", high = "#D13B2E",
      midpoint = max(hm_gg$absorb_ratio, na.rm = TRUE) / 2
    ) +
    labs(
      title = "Top differential metabolites (ggplot fallback)",
      x = "Top 20 absorb form",
      y = NULL,
      fill = "Absorb ratio"
    ) +
    theme_bw(base_size = 8) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
      panel.grid = element_blank()
    )
}
```

## 5. Session info

```{r session-info}
sessionInfo()
```
