---
title: "TRACE_Demo"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
vignette: "%\\VignetteIndexEntry{TRACE_Demo} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Data set preparation

### 1.1 load data set

Here is a data set with 15 samples grouped as:

- $^{12}\text{C}$ and $^{14}\text{N}$
- $^{12}\text{C}$ and $^{14}\text{N}$
- $^{13}\text{C}$ and $^{15}\text{N}$
- $^{12}\text{C}$ and $^{15}\text{N}$
- Blank

Use `MSdev` to load the raw data and convert it to `mzML` files.

```{r}
library(MSdev)
demo.dir <- "d:/data/PAVE_demo/rawdata"
trace.demo <- MSdev( rawDataDir =  demo.dir)
trace.demo <- MSdev_msConvert(trace.demo)
trace.demo@sampleInfo
```

### 1.2 Peaks extraction by xcms

`MSdev_xcmsProcessing` call `xcms::findChromPeaks` and `xcms::groupChromPeaks` to extract peaks and group peaks.

Use `MSdev_checkSampleInfo` to input the sample information: sample.group

Use `MSdev_set_param` to set the parameters for `xcms::findChromPeaks` and `xcms::groupChromPeaks` and then run `MSdev_xcmsProcessing` to extract peaks and group peaks.

```{r}
if(F){
  trace.demo <- MSdev_set_param(
    trace.demo,
    findChromPeaks =
      xcms::CentWaveParam(
        ppm = 5,
        prefilter = c(3,1000),
        peakwidth = c(10,30),
        snthresh = 10,
        fitgauss = T),
    groupChromPeaks =
      xcms::PeakDensityParam(
        sampleGroups = "A",
        minFraction = 0.6,
        binSize = 0.002,
        bw = 5,
        ppm = 5))
  trace.demo <- MSdev_checkSampleInfo(trace.demo)
  trace.demo <- MSdev_xcmsProcessing(trace.demo)
  
}
```

Save the processed data set to `Rdata` file, so that you can load the processed data set later without running the `MSdev_xcmsProcessing` again. Load the processed data set from `Rdata` file:

```{r}
#MSdev_save(trace.demo)

trace.demo <- MSdev_load("d:/data/PAVE_demo/MSdev_2026_06_01.Rdata")
trace.demo
```

## 2. TRACE Core

### 2.1 find CN labeling

```{r}
trace.demo <- TRACE_get_CN_net(trace.demo,0,ratio.adjust = c(1.0842,0.7641,1.0566,1.0707))
trace.demo <- TRACE_dynamic_filter(trace.demo,0)
trace.demo <- TRACE_network_assignment(trace.demo,0)
trace.demo <- TRACE_annotate(trace.demo,i.pol = 0)

trace.demo <- TRACE_get_CN_net(trace.demo,1,ratio.adjust = c(1.0842,0.7641,1.0566,1.0707))
trace.demo <- TRACE_dynamic_filter(trace.demo,1)
trace.demo <- TRACE_network_assignment(trace.demo,1)
trace.demo <- TRACE_annotate(trace.demo,i.pol = 1)

trace.demo <- TRACE_CN_labelling_ratio_adjust(trace.demo,eval_top = 0.3,plot = T)

```

## 3. TRACE paramers sensitivity

TRACE utilize xcms to extract peaks and group peaks, so the parameters of xcms will affect the TRACE result. \`xcms::CentWaveParam()' was usually used for peak detection

```{r}
xcms::CentWaveParam()
```

### 3.1 xcms parameters sensitivity

#### 3.1.1 set parameters grid and run TRACE

we set parameters grid as follows:

```{r}
param.list <- list(
  cw1 = xcms::CentWaveParam(ppm = 5, prefilter = c(3,1000), peakwidth = c(10,30), snthresh = 10),
  cw2 = xcms::CentWaveParam(ppm = 5, prefilter = c(3,1000), peakwidth = c(10,30), snthresh = 100),
  cw3 = xcms::CentWaveParam(ppm = 10, prefilter = c(3,1000), peakwidth = c(10,30), snthresh = 10),
  cw4 = xcms::CentWaveParam(ppm = 20, prefilter = c(3,1000), peakwidth = c(10,30), snthresh = 10),
  cw5 = xcms::CentWaveParam(ppm = 5, prefilter = c(5,1000), peakwidth = c(10,30), snthresh = 10))


for(i in 1:length(param.list)){
  trace.demo <- MSdev_set_param(trace.demo, findChromPeaks = param.list[[i]],
                                groupChromPeaks = xcms::PeakDensityParam(
                                  sampleGroups = "A",
                                  minFraction = 0.6,
                                  binSize = 0.002,
                                  bw = 5,
                                  ppm = param.list[[i]]@ppm ))
  trace.demo <- MSdev_xcmsProcessing(trace.demo)
  #trace.demo <- TRACE_workflow(trace.demo)
  save.name <- paste0("TRACE_xcms_sensitivity_ppm_",param.list[[i]]@ppm,"_sn_",param.list[[i]]@snthresh,"_prefilter_",param.list[[i]]@prefilter[1],"_",param.list[[i]]@prefilter[2],".Rdata")
  save.file <- paste0("d:/data/2025.12.26.PAVE2/xcms_sensitivity/",save.name)
  MSdev_save(trace.demo,save.file)
}

for(i in 1:length(param.list)){
  message_with_time()
  save.name <- paste0("TRACE_xcms_sensitivity_ppm_",param.list[[i]]@ppm,"_sn_",param.list[[i]]@snthresh,"_prefilter_",param.list[[i]]@prefilter[1],"_",param.list[[i]]@prefilter[2],".Rdata")
  save.file <- paste0("d:/data/2025.12.26.PAVE2/xcms_sensitivity/",save.name)
  obj <- MSdev_load(save.file)
  obj@advancedAna <- list()
  obj <- TRACE_workflow(obj,ratio.reconstruct = F)
  MSdev_save(obj,save.file)
}
```

#### 3.1.2 result

Plot the result under parameters grid:

```{r}
library(dplyr)
library(ggplot2)
library(MSdev)

sens.dir <- "d:/data/2025.12.26.PAVE2/xcms_sensitivity/"
save.names <- vapply(param.list, function(p) {
  paste0(
    "TRACE_xcms_sensitivity_ppm_", p@ppm,
    "_sn_", p@snthresh,
    "_prefilter_", p@prefilter[1], "_", p@prefilter[2],
    ".Rdata"
  )
}, character(1))
param.labels <- c("benchmark", "sn=100", "ppm=10", "ppm=20", "prefilter=5")

trace.all <- lapply(seq_along(param.list), function(i) {
  obj <- MSdev_load(file.path(sens.dir, save.names[i]))
  trace.neg <- obj@advancedAna$TRACE$Negative
  trace.pos <- obj@advancedAna$TRACE$Positive
  dplyr::bind_rows(
    if (!is.null(trace.neg) && nrow(trace.neg)) {
      trace.neg %>% dplyr::mutate(polarity = "Negative")
    },
    if (!is.null(trace.pos) && nrow(trace.pos)) {
      trace.pos %>% dplyr::mutate(polarity = "Positive")
    }
  ) %>%
    dplyr::mutate(
      param = names(param.list)[i],
      param.label = param.labels[i]
    )
})
trace.all <- dplyr::bind_rows(trace.all)

type.levels <- c("metabolite", "adduct", "fragment", "isotope", "unknown")
type.comp <- trace.all %>%
  dplyr::mutate(
    type = dplyr::case_when(
      is.na(type) | type == "" ~ "unknown",
      TRUE ~ as.character(type)
    ),
    type = factor(type, levels = type.levels),
    param = factor(param, levels = names(param.list)),
    param.label = factor(param.label, levels = param.labels)
  ) %>%
  dplyr::count(param, param.label, type, name = "n") %>%
  dplyr::ungroup()

p.comp <- ggplot2::ggplot(type.comp, ggplot2::aes(x = param.label, y = n, fill = type)) +
  ggplot2::geom_col(position = "stack", width = 0.7) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
  ggplot2::scale_fill_brewer(palette = "Set2") +
  ggplot2::labs(
    x = NULL,
    y = "Number of features",
    fill = "Annotation type",
    title = "TRACE annotation composition across xcms parameter grid"
  ) +
  ggplot2::theme_bw(base_size = 6) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(size = 4),
    legend.position = "right"
  )

print(p.comp)

compound.sets <- lapply(seq_along(param.list), function(i) {
  trace.all %>%
    dplyr::filter(param == names(param.list)[i]) %>%
    dplyr::filter(type == "metabolite", !is.na(compound_id)) %>%
    dplyr::pull(compound_id) %>%
    unique()
})
names(compound.sets) <- param.labels

if (requireNamespace("UpSetR", quietly = TRUE)) {
  compound.mat <- UpSetR::fromList(compound.sets)
  p2 <- UpSetR::upset(
    compound.mat,
    nsets = length(compound.sets),
    order.by = "freq",
    decreasing = TRUE,
    mainbar.y.label = "Compound intersection size",
    sets.x.label = "Compounds per parameter set",
    text.scale = c(1.3, 1.1, 1, 1, 1.2, 1),
    main.bar.color = "#66C2A5",
    sets.bar.color = "#FC8D62"
  )
} else {
  message(
    "Install UpSetR to draw the compound_id UpSet plot: ",
    "install.packages('UpSetR')"
  )
}
```

### 3.2 PAVE cor sensitivity

`TRACE_get_CN_net()` uses `TRACE_cor_cutoff` to filter CN hits with correlation score below the cutoff.

#### 3.2.1 set parameters grid and run TRACE

We test a grid of `TRACE_cor_cutoff` values on the same xcms-processed data set:

```{r}
cor.list <- seq(0.5,0.95,0.05)
cor.labels <- paste0("cor=", cor.list)
ratio.adj <- c(1.0842, 0.7641, 1.0566, 1.0707)
base.file <- "d:/data/PAVE_demo/MSdev_2026_06_01.Rdata"
sens.dir <- "d:/data/2025.12.26.PAVE2/cor_sensitivity/"
dir.create(sens.dir, recursive = TRUE, showWarnings = FALSE)

for (i in seq_along(cor.list)) {
  message_with_time()
  obj <- MSdev_load(base.file)
  obj@advancedAna <- list()
  cor.cut <- cor.list[i]
  for (i.pol in 0:1) {
    obj <- TRACE_get_CN_net(
      obj,
      i.pol = i.pol,
      ratio.adjust = ratio.adj,
      TRACE_cor_cutoff = cor.cut
    )
  }
  save.name <- paste0("TRACE_cor_sensitivity_cutoff_", cor.cut, ".Rdata")
  save.file <- file.path(sens.dir, save.name)
  MSdev_save(obj, save.file)
}
```

#### 3.2.2 result

Plot the result under `TRACE_cor_cutoff` grid:

```{r}
library(dplyr)
library(ggplot2)
library(MSdev)

save.names <- paste0("TRACE_cor_sensitivity_cutoff_", cor.list, ".Rdata")

cn.hit.all <- lapply(seq_along(cor.list), function(i) {
  obj <- MSdev_load(file.path(sens.dir, save.names[i]))
  cn.neg <- obj@advancedAna$TRACE_temp$Negative$cn.net.hit
  cn.pos <- obj@advancedAna$TRACE_temp$Positive$cn.net.hit
  dplyr::bind_rows(
    if (!is.null(cn.neg) && nrow(cn.neg)) {
      cn.neg %>% dplyr::mutate(polarity = "Negative")
    },
    if (!is.null(cn.pos) && nrow(cn.pos)) {
      cn.pos %>% dplyr::mutate(polarity = "Positive")
    }
  ) %>%
    dplyr::mutate(
      cor.cut = cor.list[i],
      param.label = cor.labels[i]
    )
})
cn.hit.all <- dplyr::bind_rows(cn.hit.all)

cn.seed <- cn.hit.all %>%
  dplyr::group_by(cor.cut, param.label, from) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup()

formula.comp <- cn.seed %>%
  dplyr::mutate(
    TRACE_formula = dplyr::coalesce(as.character(TRACE_formula), "unknown"),
    cor.cut = factor(cor.cut, levels = cor.list),
    param.label = factor(param.label, levels = cor.labels)
  ) %>%
  dplyr::count(cor.cut, param.label, TRACE_formula, name = "n") %>%
  dplyr::ungroup()

p.comp <- ggplot2::ggplot(formula.comp, ggplot2::aes(x = param.label, y = n, fill = TRACE_formula)) +
  ggplot2::geom_col(position = "stack", width = 0.7) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
  ggplot2::scale_fill_brewer(palette = "Set2") +
  ggplot2::labs(
    x = NULL,
    y = "Number of CN seeds",
    fill = "CN formula",
    title = "CN seed composition across TRACE_cor_cutoff grid"
  ) +
  ggplot2::theme_bw(base_size = 6) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(size = 4, angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p.comp)

seed.sets <- lapply(seq_along(cor.list), function(i) {
  cn.seed %>%
    dplyr::filter(cor.cut == cor.list[i]) %>%
    dplyr::pull(from) %>%
    unique()
})
names(seed.sets) <- cor.labels

if (requireNamespace("UpSetR", quietly = TRUE)) {
  seed.mat <- UpSetR::fromList(seed.sets)
  UpSetR::upset(
    seed.mat,
    nsets = length(seed.sets),
    order.by = "freq",
    decreasing = TRUE,
    mainbar.y.label = "CN seed intersection size",
    sets.x.label = "CN seeds per TRACE_cor_cutoff",
    text.scale = c(1.3, 1.1, 1, 1, 1.2, 1),
    main.bar.color = "#66C2A5",
    sets.bar.color = "#FC8D62"
  )
} else {
  message(
    "Install UpSetR to draw the CN seed UpSet plot: ",
    "install.packages('UpSetR')"
  )
}
```

## 4. TRACE evaluation

### 4.1 RSD

show the RSD of features in the 4 groups of samples: S12C14N, S12C15N, S13C14N, S13C15N

```{r}

plot_TRACE_RSD(
  trace.demo, 
  i.pol = 1 )


```

### 
