---
title: "TRACE"
output: html_notebook
vignette: >
  %\VignetteIndexEntry{TRACE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# TRACE refactor
1. read the code of TRACE.R, refactor to three parts, each part is a function:
  - TRACE_get_CN_net
  - TRACE_dynamic_filter
  - TRACE_network_assignment
2. write a new function TRACE_workflow to call the three functions above to complete the TRACE analysis.
3. keep the original code for reference, all the return should be a MSdev object. data required for the following
analysis should be stored in the MSdev object@advancedAna$TRACE_temp
4. some of function in TRACE() is called from pacakge MSdev, search for the function in MSdev, if it is exported, add MSdev:: to the function name. if not, notice me to export it from MSdev.

```{r}
# New refactored workflow entry point
# object <- TRACE_workflow(object)

# Optional step-by-step execution for one polarity
# object <- TRACE_get_CN_net(object, i.pol = 0)
# object <- TRACE_dynamic_filter(object, i.pol = 0)
# object <- TRACE_network_assignment(object, i.pol = 0)
# object <- TRACE_annotate(object, i.pol = 0)
```


## TRACE_CN_labelling_ratio_adjust
1. add a function TRACE_get_CN_labelling_ratio to get the CN labelling ratio from the MSdev object:
  - input: MSdev object
  - this function check all the CN_hit from 'TRACE_get_CN_net'
  - for each CN_hit, it will check the labelling ratio of the CN_hit in the 4 groups of samples: S12C14N, S12C15N, S13C14N, S13C15N
  - extract the labelling ratio of CN_hit will be stored in a data.frame with columns:
    - TRACE_seed: the feature_id of the CN_hit
    - TRACE_formula: the formula of the CN_hit
    - TRACE_cor: the correlation score of the CN_hit
    - S12C14N: the labelling ratio of the CN_hit in the S12C14N group
    - S12C15N: the labelling ratio of the CN_hit in the S12C15N group
    - S13C14N: the labelling ratio of the CN_hit in the S13C14N group
    - S13C15N: the labelling ratio of the CN_hit in the S13C15N group
2. TRACE_CN_labelling_ratio_adjust should extract ratio of corresponding feature in the 4 groups of samples: 
  - S12C14N should extract C0N0 feature, 
  - S12C15N should extract C0NY feature, 
  - S13C14N should extract CXN0 feature, 
  - S13C15N should extract CXNY feature
```{r}
# Run TRACE steps first, then collect CN seed labelling ratio:
# object <- TRACE_workflow(object)
# cn_ratio <- TRACE_get_CN_labelling_ratio(object)
# head(cn_ratio)
```