---
title: "TRACE"
output: html_notebook
vignette: >
  %\VignetteIndexEntry{TRACE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# TRACE refactor
1. read the code of TRACE.R, refactor to three parts, each part is a function:
  - TRACE_get_CN_net
  - TRACE_dynamic_filter
  - TRACE_network_assignment
2. write a new function TRACE_workflow to call the three functions above to complete the TRACE analysis.
3. keep the original code for reference, all the return should be a MSdev object. data required for the following
analysis should be stored in the MSdev object@advancedAna$TRACE_temp
4. some of function in TRACE() is called from pacakge MSdev, search for the function in MSdev, if it is exported, add MSdev:: to the function name. if not, notice me to export it from MSdev.

```{r}
# New refactored workflow entry point
# object <- TRACE_workflow(object)

# Optional step-by-step execution for one polarity
# object <- TRACE_get_CN_net(object, i.pol = 0)
# object <- TRACE_dynamic_filter(object, i.pol = 0)
# object <- TRACE_network_assignment(object, i.pol = 0)
# object <- TRACE_annotate(object, i.pol = 0)
```


## TRACE_CN_labelling_ratio_adjust
1. add a function TRACE_get_CN_labelling_ratio to get the CN labelling ratio from the MSdev object:
  - input: MSdev object
  - this function check all the CN_hit from 'TRACE_get_CN_net'
  - for each CN_hit, it will check the labelling ratio of the CN_hit in the 4 groups of samples: S12C14N, S12C15N, S13C14N, S13C15N
  - extract the labelling ratio of CN_hit will be stored in a data.frame with columns:
    - TRACE_seed: the feature_id of the CN_hit
    - TRACE_formula: the formula of the CN_hit
    - TRACE_cor: the correlation score of the CN_hit
    - S12C14N: the labelling ratio of the CN_hit in the S12C14N group
    - S12C15N: the labelling ratio of the CN_hit in the S12C15N group
    - S13C14N: the labelling ratio of the CN_hit in the S13C14N group
    - S13C15N: the labelling ratio of the CN_hit in the S13C15N group
2. TRACE_CN_labelling_ratio_adjust should extract ratio of corresponding feature in the 4 groups of samples: 
  - S12C14N should extract C0N0 feature, 
  - S12C15N should extract C0NY feature, 
  - S13C14N should extract CXN0 feature, 
  - S13C15N should extract CXNY feature
3. the ratio should be normalized to the average of C0N0(S12C14N), not the sum of the 4 groups of samples.
4. when the formula is CXN0, the C0NY should be equal to C0N0, the CXNY should be equal to CXN0, not NA
5. add an argument 'eval_top', slice the top prop of eval_top TRACE_cor of the TRACE_CN_labelling_ratio to evaluate the labelling ratio. calculate the median of the labelling ratio of each group using the top eval_top TRACE_cor. message: 'Evaluate the adjusted labelling ratio with top xx% TRACE_cor, run TRACE_get_CN_net(ratio.adjust = c(A,B,C,D)) to get the best CN label.'
6. add argument 'plot', if TRUE, plot the labelling ratio of each group using ggplot2
7. the result of TRACE_CN_labelling_ratio_adjust should be stored in the MSdev object@advancedAna$TRACE_temp$cn.group.ratio
8. add argument 'reconstruct' to the function, if TRUE, reconstruct the CN_hit with the labelling ratio of the 4 groups of samples.
9. rename 'TRACE_get_CN_labelling_ratio' to 'get_TRACE_CN_labelling_ratio', 'TRACE_CN_labelling_ratio_adjust' also store the ratio.df
10. integrate the function into the TRACE_workflow, call the function in the TRACE_workflow after 'TRACE_get_CN_net'


## TRACE_get_CN_net
1. add argument 'ratio.adjust' to the function, the default value is c(1,1,1,1), this ratio was used to compare pattern of CN features, and pass into 'get_ideal_CN_ratio' to get the theoretical labelling ratio of the CN_hit in the 4 groups of samples.
2. add argument 'TRACE_cor_cutoff' to the "TRACE_get_CN_net", the default value is 0.7, this cutoff was used to filter the CN_hit with TRACE_cor less than TRACE_cor_cutoff.


## TRACE_export
1. write a function TRACE_export to export the TRACE result to a xlsx file.

## plot_TRACE_labeling_fraction 
1. write a function 'plot_TRACE_labeling_fraction' to plot the labeling fraction of features:
  - input: MSdev object
  - this function plot the labeling fraction of features for each metabolties and sample:
    - for metabolites, all possible CxNy should be count: feature matched to C0N0, C1N0, C1N2, ..., CxNy should be count
    - the ratio lableing ratio for samples:
    - S13C15N: CxNy/total 
    - S12C15N: C0Ny/total
    - S13C14N: CxN0/total
    - S12C14N: C0N0/total
  - use geom_box and geom_point to show lableing ratio
2. add label to indicate every group's calculation, such as  S13C15N: CxNy/total 
3. add p2: geom_density to show the distribution of the labeling ratio, use patchwork to combine the two plots.